---
title: "comparison_catalyst_data"
author: "Qi Miao"
output: html_document
---

#Code adapted from correlation_analysis.Rmd Vito Zanotelli, et al. https://github.com/BodenmillerGroup/cyTOFcompensation/blob/master/scripts/correlation_analysis.Rmd

```{r Libraries}
library(CytoSpill)

library(flowCore)
library(Rtsne)
library(Rphenograph)
library(CATALYST)
library(dplyr)
library(data.table)
library(dtplyr)
library(RColorBrewer)
library(gplots)
library(cba)
library(Hmisc)
library(mclust)
```

###  Define some helper functions
```{r Define some helper functions functions}
flowFrame2dt <- function (datFCS){
  # converts a flow frame to a data table
  dt = flowCore::exprs(datFCS)
  colnames(dt) = make.names(make.unique(colnames(dt)))
  fil = !is.na(datFCS@parameters@data$desc)
  colnames(dt[, fil]) <- datFCS@parameters@data$desc[fil]
  dt <- data.table::data.table(dt)
  uninames = make.names(make.unique(datFCS@parameters@data$name))
  uni_newnames = make.names(make.unique(datFCS@parameters@data$desc))
  data.table::setnames(dt, uninames[fil], uni_newnames[fil])
  return(dt)
}

censor_dat <- function (x, quant = 0.999){
  # censors the upper limit vector's value using the provided quantile
  q = quantile(x, quant)
  x[x > q] = q
  return(x)
}

```


### Set the configuration variables
```{r Files}
# the FCS file used
fn_cells = '../data/160805_Exp3_cells-only.fcs'
name_condition = 'Exp3'

# the file containing the spillover matirx generated by CATALYST 
fn_sm = c('../data/160805_Exp3_spillover_matrix.csv')

folder_out = '../data/CATALYST_comparison/'
do_load = T

# subsample cellnumber
nsamp = 20000

transf = function(x) asinh(x/5)

# defines some channel names that should not be considered
bad_channels = c(  "CD3"   ,  "CD45","SampleID","Time", "Event_length", "MCB102", "MCB104", "MCB105", "MCB106",
                   "MCB108","MCB110","MCB113","MCB115" ,"Beads140" , "File.Number",
                   "DNA193",    "DNA191" ,   "Live194",      "Live195" ,     "beadDist",     "barcode",
                   "89Y"  ,  "102Pd_BC1" , "104Pd_BC2" ,"105Pd_BC3" ,"106Pd_BC4","108Pd_BC5", "110Pd_BC6" ,    
                   "113In_BC7", "115In_BC8", "138Ba","140Ce", "191Ir_DNA1" , 
                   "193Ir_DNA2","195Pt","196Pt","208Pb", "Center","Offset","Width",          
                   "Residual" ,  "102Pd"  , "103Rh" , "104Pd" , "105Pd" , "106Pd","108Pd",
                   "110Pd", "113In","115In" ,"157Gd" ,"155Gd-CD273",'File-Number'
)

col_list  <- c("#DC050C", "#FB8072", "#1965B0", "#7BAFDE", "#882E72", 
               "#B17BA6", "#FF7F00", "#FDB462", "#E7298A", "#E78AC3", 
               "#33A02C", "#B2DF8A", "#55A1B1", "#8DD3C7", "#A6761D", 
               "#E6AB02", "#7570B3", "#BEAED4", "#666666", "#999999", 
               "#aa8282", "#d4b7b7", "#8600bf", "#ba5ce3", "#808000", 
               "#aeae5c", "#1e90ff", "#00bfff", "#56ff0d", "#ffff00")
```

### Read the data as a flow frame
```{r Load the single cell data}
ff <- flowCore::read.FCS(fn_cells)
```

### Subsamples to speed the calculations up
```{r Subsample the data to speed the calculatios up}
set.seed(1234)
if (!is.na(nsamp)){
  ff = ff[sample(nrow(ff), nsamp)]
}
```

### Load the spillover matrixes generated and saved by CATALYST
```{r Load SM}
dat_sms = lapply(fn_sm, function(fn) read.csv(fn, row.names = 1))
names(dat_sms) = sapply(fn_sm, function(fn) basename(fn))
```

### Load the spillover matrix generated by CytoSpill
```{r}
es <- flowCore::exprs(ff)
###only column 15 to 51 were used as the same as CATALYST
# spillmat_results <- GetSpillMat(es, cols=15:51)
set.seed(2)
spillmat_cutoffs <- CytoSpill::.DeriveCutoffs(es, cols=15:51, n = nrow(es), flexrep = 30)
model <- .EstimateSpill(data=es, cutoffs=spillmat_cutoffs, cols=15:51, upperbound = 0.1, neighbor = 1)
threshold = 0.1
estimates <- model[[1]]
xcols <- model[[2]]
spillmat <- diag(length(xcols))
for (i in 1:length(xcols)) {
  if (!is.na(xcols[[i]][1])) {
    for (j in 1:length(xcols[[i]])) {
      if (!is.na(estimates[[i]][j])){
        spillmat[xcols[[i]][j],i] <- min(estimates[[i]][j],threshold)
       }
    }
  }
}

sm_cytospill <- dat_sms
#substitute spill matrix values with spill matrix generated with CytoSpill
sm_cytospill[[1]][15:51,15:51] <- spillmat
names(sm_cytospill) <- "cytospill_estimated_spillover_matrix"
```

```{r}
#list of two spillover matrix one from single label control, one from CytoSpill
dat_sms[[2]] <- sm_cytospill[[1]]
names(dat_sms)[2] <- "cytospill_estimated_spillover_matrix"
```

### some notes here!
here the catalyst spillover matrix has all the colunms, not only the three sources.
### Compensate the data

```{r Compensate the data using all the compensation matrices}
dats_compmeta = data.table(compensation= c('raw', paste(names(dat_sms),'_nnls')))
dats_compmeta[, is_raw := compensation == 'raw']
ff_list = list()
ff_list[[1]] = ff

#source modified scripts from CATALSYT package, since the original version has error occurs.
source("../scripts/adaptSpillmat_(remove check_sm).R")
source("../scripts/helpers-compensation.R")

for (i in seq_along(dat_sms)){
  tsm = dat_sms[[i]]
  # helps with numerical instabilities sometimes observed during saving/loading
  tsm = round(tsm, digits=8)
  ff_list[[i+1]] = CATALYST::compCytof(ff, tsm, method='nnls')
}

# for (i in seq_along(dat_sms)){
#   tsm = dat_sms[[i]]
#   # There seem some numerical instabilities when loading the data (e.g.  -5.596894e-33). 
#   # rounding fixes this issue.
#   tsm = round(tsm, digits=8)
#   ff_list[[i+(length(dat_sms)+1)]] = CATALYST::compCytof(ff, tsm, method='nnls')
# }

names(ff_list) <- dats_compmeta$compensation
# ff_list with raw, flow compensated, nnls compensated
```

### Tidy the data into data.tables

```{r Make a tidy data.table out of the data}
tidy_ff <- function(x, nm_condition){
  x = flowFrame2dt(x)
  x[, id:=paste(as.character(1:.N), nm_condition)]
  x =  melt.data.table(x, id.vars = 'id',variable.name = 'channel', value.name='counts', variable.factor = F)
  x[, counts_transf:= transf(counts)]
  
  setkey(x, id)
  
  return(x)
}

dats= lapply(ff_list, function(x) tidy_ff(x, name_condition))
```

### Clean up some channel names
```{r Clean channel names}
clean_channels = function(x){
  x[, channel :=as.character(gsub('X','',as.character(channel))) , by=channel]
  x[, channel :=gsub('\\.','-',as.character(.BY)) , by=channel]
  x[, channel :=gsub(' ','-',as.character(.BY)) , by=channel]
  return(x)
}
dats= lapply(dats, clean_channels)
```


### Create a metadata table
```{r Metadata}

# for additional, global cell specific information, e.g. condition
datcell = dats[['raw']] %>%
  dplyr::select(-c(counts, counts_transf, channel)) %>%
  dplyr::filter(!duplicated(id)) %>%
  mutate(condition= name_condition)
datcell <- as.data.table(datcell)
setkey(datcell, id)

datmeta =  data.table(name=ff@parameters$name, channel=ff@parameters$desc)
datmeta = clean_channels(datmeta)
setkey(datmeta, channel)

good_channels =unique(datmeta$channel)[!unique(datmeta$channel) %in% bad_channels]

```

### Add information which channels contain antibodies against the same target
```{r Add grouping antibody information}
datmeta[,antibody:=gsub('.....-','',channel)]

#datmeta[antibody=='CD8b',antibody:='CD8']
datmeta[,n_chan := length(unique(channel)), by=antibody]
datmeta[,metal:=substr(channel, 4, 5)]
datmeta[metal =='', metal:= as.character(1:.N)]
```

## Clustering and dimensionality reduction using TSNE
### Cluster the data using phenograph
```{r Define a helper function to do phenograph with the data structure used}

# adapted from https://github.com/lmweber/FlowSOM-Rtsne-example/blob/master/FlowSOM_Rtsne_example.R

do_phenograph<- function(data, channels, valuevar= 'counts_transf', seed=1234, subsample=FALSE){
  # A helper function
  
  pheno_dat = data.table::dcast.data.table(data[channel %in% channels, ],id~channel,value.var = valuevar)
  all_ids = pheno_dat$id
  
  if (subsample == FALSE){
    subsample=1
  } 
  
  
  
  sampids = pheno_dat[, sample(id, floor(.N*subsample),replace = F)]
  pheno_dat_samp = pheno_dat[id %in%sampids, ]
  ids =pheno_dat_samp$id
  pheno_dat_samp[, id:=NULL]
  set.seed(seed)
  rpheno_out = Rphenograph::Rphenograph(pheno_dat_samp)
  cluster = igraph::membership(rpheno_out[[2]])
  pheno_clust = data.table::data.table(x=ids)
  setnames(pheno_clust, 'x', 'id')
  pheno_clust[, cluster:=factor(cluster[as.character(seq(length(ids)))])]
  data.table::setkey(pheno_clust, 'id')
  pheno_clust = pheno_clust[all_ids ,]
  
  return(pheno_clust)
}
```

```{r Run phenograph on the data before and after compensation}
datspheno= lapply(dats, function(dat) do_phenograph(dat, good_channels, valuevar = 'counts_transf', seed=1234))
```

### calculate tsne maps

```{r}
calc_tsne <- function (input_dat, channels, value_var = "counts", channel_var = "channel", 
                       id_var = "id",  scale = F, verbose = T, dims = 2, seed=1234,...){
  set.seed(seed)
  ids = input_dat[!duplicated(get(id_var)), get(id_var)]
  dt = data.table::dcast(input_dat[(get(channel_var) %in% 
                                      channels) & get(id_var) %in% ids], formula = as.formula(paste(id_var, 
                                                                                                    "~", channel_var)), value.var = value_var)
  if (scale) {
    tsnedat = scale(dt[, channels, with = F])
  }
  else {
    tsnedat = dt[, channels, with = F]
  }
  tsne_out <- Rtsne::Rtsne(tsnedat, verbose = verbose, dims = dims, 
                           ...)
  tsne_out$Y = data.table(tsne_out$Y)
  setnames(tsne_out$Y, names(tsne_out$Y), paste("bh", names(tsne_out$Y), 
                                                sep = "_"))
  tsne_out$Y$id = dt[, get(id_var)]
  data.table::setnames(tsne_out$Y, "id", id_var)
  data.table::setkeyv(tsne_out$Y, id_var)
  tsne_out$channels = channels
  tsne_out$scale = F
  return(tsne_out)
}

# Setup some colors for plotting
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))

```

The TSNE map is calculated on the uncompensated values
```{r Calculate the tsne map}
ttsne =  calc_tsne(dats[['raw']], good_channels, value_var = 'counts_transf', channel_var = "channel")
```

### The phenograph clusters are plotted on the TSNE map


```{r Plot the phenograph cluster on the tsne map}

tclust = datspheno[['raw']]
tdat = dats[['raw']]

p = ggplot(subset(tdat, !duplicated(id))[ttsne$Y][tclust], aes(x=bh_V1, y=bh_V2))+
  geom_point(size=0.3, alpha=1, aes(color=as.factor(cluster)))+
  scale_color_manual(values = col_list)+
  ggtitle('Uncompensated data tsne with Phenograph clusters')+
  guides(color=guide_legend(override.aes=list(size=5), ncol=2))+
  theme(strip.background = element_blank(),
        panel.background=element_rect(fill='white', colour = 'black'),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        plot.background=element_blank(),
        legend.key = element_blank())
p

```

### select phenograph clustering to be used for all downstream analysis
The phenograph based on the uncompensated data is used for all future plots

```{r select the phenograph clustering}
clusterdat = datspheno[['raw']]
```


### Map the markers on the tsne map

In the following plots the asinh(x/5) transformed intensities were normalized between 0-1 by using the 0.99 percentile of the data.

```{r, fig.height=6, fig.width=15}
ptsne = ttsne
pdats = dats

for (tdatname in names(pdats)){
  tdat = pdats[[tdatname]]
  tdat[, c_counts := censor_dat(counts_transf,0.99), by=channel]
  tdat[, c_counts_scaled := c_counts, by=channel]
  tdat[, c_counts_scaled := (c_counts_scaled/max(c_counts_scaled)), by=channel]
  p = ggplot(subset(tdat, channel %in% good_channels)[ptsne$Y],
             aes(x=bh_V1, y=bh_V2, color=c_counts_scaled))+
    facet_wrap(~channel, scales = "free", ncol = 9)+
    geom_point(alpha=0.5, size=0.3)+
    scale_color_gradientn(colours=rev(brewer.pal(11, 'Spectral')), name='Counts', limits=c(0, 1))+
    ggtitle(tdatname)+
    theme(strip.background = element_blank(),
          strip.text.x = element_text(size = 11),
          axis.line=element_blank(),
          axis.text.x=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),
          panel.background=element_blank(),
          panel.border=element_blank(),
          panel.grid.major=element_blank(),
          panel.grid.minor=element_blank(),
          plot.background=element_blank()) 
  print(p)
  #ggsave(filename = file.path(folder_out, paste(tdatname, 'tsne_all_markers.png',sep='_')), plot = p,width=15, height=6, dpi = 300)
  
}

```



## Plot heatmaps of the phenograph cluster statistics

The chunck below defines some convenience functions
```{r Helper functions}
get_summarydat <- function(ssdat, clusterdat, valuevar){
  # calculates summary statistics per cluster
  summary_dat = ssdat[clusterdat][ ,list(
    median_val = median(get(valuevar), na.rm=T),
    mean_val = median(get(valuevar), na.rm=T),
    cell_cluster=.N
  ), by=.(channel,cluster)]
  return(summary_dat)
}

dat2mat <- function(data, formula, valuevar){
  # calculates a matrix suitable for a heatmap from the long data
  hm_dat = dcast.data.table(data =data, formula =formula,
                            value.var = valuevar)
  
  trownames = hm_dat$cluster
  
  # convert to a matrix
  hm_dat = as.matrix(hm_dat[,-1,with=F])
  row.names(hm_dat) = trownames
  
  return(hm_dat)
}


```

### Heatmap of the cluster medians

For each phenograph cluster the cluster median is calculated and a heatmap is generated.
The all heatmap are clustered based on the uncompensated data in order to make them visually easily comparable.
The colorscale corresponds to the asinh(x/5) transformed count and is kept constant for all the plots to make them directly comparable.

```{r}
firstclust = 1
pdats = dats
par(cex.main = 1)
names(pdats) = c("Uncompensated", "Single-stained controls compensated", "CytoSpill compensated")
for (i in seq_along(pdats)){
  cur_name = names(pdats)[i]
  hm_dat = pdats[[i]][channel %in% good_channels,] %>%
    get_summarydat(clusterdat, 'counts_transf') %>%
    dat2mat('cluster ~ channel','median_val')
  
  if (firstclust){
    firstclust=F
    tdist = as.dist(1-cor(t(hm_dat), method="pearson"))
    tdist[is.na(tdist)] = 0
    hr <- hclust(tdist, method="ward.D")
    co_r <- order.optimal(tdist, hr$merge)
    hr$merge = co_r$merge
    hr$order = co_r$order
    
    tdist = as.dist(1-cor((hm_dat), method="pearson"))
    tdist[is.na(tdist)] = 0
    hc <- hclust(tdist, method="ward.D")
    co_c <- order.optimal(tdist, hc$merge)
    hc$merge = co_c$merge
    hc$order = co_c$order
  }
  if (any(hm_dat<0)){
    cols =  rev(brewer.pal(11,'RdBu'))
    cmap = colorRampPalette(cols)
    
  } else {
    cols =  rev(brewer.pal(11,'RdBu'))
    cmap = colorRampPalette(cols[6:11])
    
  }
  
  heatmap.2(hm_dat,
            scale ='none',
            trace = "none",
            # col=cmap(75),
            col=rev(brewer.pal(11,'RdBu')),
            Rowv=as.dendrogram(hr),
            #RowSideColors=sel_col_vector[group_levels],
            #dendrogram='column',
            Colv=as.dendrogram(hc),
            density.info ='none',
            #Colv =NA,
            #keyorient=2,                                                  
            #sepwidth = c(0,0),
            cexRow=0.9,
            cexCol=0.5,
            margins=c(4,4),
            main=cur_name
            #ylab ='Genes',
            #comments = data.frame(names = row.names(tab_Prot))
  )
}


```

These heatmaps use arcsinh(x/5) scaled data. There is clearly some spillover vanishing upon compensation. After compensation the same agentibodies stained in the different channels look much more comparable.
With the 'classical' compensation a slight overcompensation is visible that is not visible an the NNLS compensation.


```{r}
sessionInfo()
```

